On Optimality of Finite Element Integration

### 2015-09-24 TOMS_Paper/assembly.tex
Reworked basically the whole perf eval section

We evaluate the speed-ups achieved by a wide variety of transformation systems over ``original code'' produced by the FEniCS Form Compiler (i.e., no optimizations applied). We analyze the following transformation systems
\begin{itemize}
\item FEniCS Form Compiler: optimized quadrature mode (work presented in~\citeN{quadrature1}). Referred to as \texttt{quad} 
\item FEniCS Form Compiler: tensor mode (work presented in~\citeN{FFC-TC}). Referred to as \texttt{tens} 
\item FEniCS Form Compiler: automatic mode (choice between \texttt{tens} and \texttt{quad} driven by heuristic, detailed in~\citeN{Fenics} and summarized in Section~\ref{sec:qualitative}). Referred to as \texttt{auto} 
\item UFLACS: a novel back-end for the FEniCS Form Compiler (whose primary goals are improved code generation time and runtime). Referred to as \texttt{ufls} 
\item COFFEE: generalized loop-invariant code motion (work presented in~\citeN{Luporini}). Referred to as \texttt{cfO1} 
\item COFFEE: optimal loop nest synthesis plus symbolic execution for zero-elimination (work of this article). Referred to as \texttt{cfO2} 
\end{itemize}

@@

We report the results of our experiments in Figures~\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity} as three-dimensional plots. The axes represent $q$, $nf$, and code transformation system. We show one subplot for each problem instance $[q, nf]$, with the code transformation system varying within each subplot. The best variant for each problem instance is given by the tallest bar, which indicates the maximum speed-up over non-transformed code.

@@

The main observation is that our transformation strategy does not always guarantee minimum execution time. In particular, 5$\%$ of the test cases (3 out of 56, without counting marginal differences) show that \texttt{cfO2} was not optimal in terms of runtime. The most significant of such test cases is the elastic model with $[q=4, nf=0]$. There are two reasons for this. Firstly, low level optimization can have a significant impact on actual performance. For example, the aggressive loop unrolling in \texttt{tens} eliminates operations on zeros and reduces the working set size by not storing entire temporaries; on the other hand, preserving the loop structure can maximize the chances of autovectorization. Secondly, memory constraints are critical, particularly the transformation strategy adpopted when exceeding $T_H$. We will later thoroughly elaborate on all these aspects.

A second observation regards the use of pre-evaluation in \texttt{cfO2}. The fact that the \texttt{cfO2} and \texttt{cfO1} bars match suggests pre-evaluation was found unprofitable and, therefore, rejected. The trend -- although there are quite a few exceptions -- is that as $nf$ increases, pre-evaluation tends not to be performed. This is a consequence of a larger increase factor, as discussed in Section~\ref{sec:op_count}.

A third point concerns autovectorization. The discretizations employed result in inner loops and basis function tables of size multiple of the machine vector length. This, combined with the chosen compilation flags, promotes autovectorization in the majority of code variants. An exception is \texttt{quad} due to the presence of indirection arrays in the generated code. In \texttt{tens}, loop nests are fully unrolled, so the standard loop vectorization is not feasible; manual inspection of the compiled code suggests, however, that block vectorization (\cite{SLP-vect}) is often triggered. In \texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces have similar structure (there are a few exceptions in \texttt{cfO2} due to zero-elimination), with loop vectorization being regularly applied, as far as we could evince from compiler reports and manual inspection of assembly code.

Other less relevant aspects are
\begin{itemize}
\item The lack of (or marginal) improvements in all code variants when $q=1$ is due to the cost of matrix insertion, which outweighs that of local assembly.
\item The slow-downs experienced with \texttt{quad} are imputable to the presence of indirection arrays in the generated code; especially with dense tables, trading autovectorization for a relatively small save in floating point operations can be counter-productive (see also Section~\ref{sec:zeros}).
\item If a bar or a subplot is missing, it means that the form compiler failed at generating code because of either exceeding the memory limit of the system or being unable to handle the form.
\end{itemize}

\paragraph{Mass}
We start with the simplest of the bilinear forms that we are investigating, the mass matrix. From analysis of Figure~\ref{fig:mass}, we see that \texttt{cfO2} shows the highest speed-ups. It is worth noting how \texttt{auto} does not select the fastest implementation (out of \texttt{quad} and \texttt{tens}) in some of the test cases: \texttt{auto} always opts for \texttt{tens}, while as $nf \geq 2$ \texttt{quad} tends to be preferable. On the other hand, one can infer that \texttt{cfO2} constantly makes correct decisions about pre-evaluation.

\paragraph{Helmholtz}
The main message of Figure~\ref{fig:helmholtz} is that \texttt{cfO2} still outperforms the other code generation systems. For $nf=0$, the adoption of pre-evaluation results in increasingly notable speed-ups over \texttt{cfO1} as $q$ increases; \texttt{tens} is comparable to \texttt{cfO2}, with \texttt{auto} making the right choice. With $nf=1$, \texttt{auto} still picks \texttt{tens}, although this time the choice is sub-optimal: this can indirectly be inferred from \texttt{cfO2}, which rejects pre-evaluation for any $q$. For $nf=2$ and $nf=3$, \texttt{auto} reverts to \texttt{quad}, which would theoretically be the right choice (the flop count is much lower than in \texttt{tens} or what would be produced by pre-evaluation); however, the generated code suffers from the presence of indirection arrays, which break autovectorization and ``traditional'' code motion. The sporadic slow-downs or marginal improvements exhibited by \texttt{ufls} are imputable to the presence of sharing.

An interesting experiment we performed was relaxing the memory threshold by setting it to $T_H = L3_{size}$. We found that this makes \texttt{cfO2} generally slower as $nf \geq 2$, with a maximum slow-down of 2.16$\times$ with $nf=2, q=2$. Presumably, the effects of not having a sensible threshold could even be worse in parallel runs, since the L3 cache is a shared memory.

\paragraph{Elasticity}
The performance results for the elastic model are displayed in Figure~\ref{fig:elasticity}. The fact that \texttt{auto} opts for \texttt{tens} when $nf=1$ leads, generally, to sub-optimal execution times. This is different than in the case of the Helmholtz equation, in which the choice of \texttt{tens} was generally correct when $nf=1$. Pre-evaluation is never triggered in \texttt{cfO2}, because of either being estimated sub-optimal or exceeding $T_H$. In the former case, the performance improvements exhibited by \texttt{cfO2} over \text{cfO1} (and, in general, over all code variants) are due to the elimination of sharing and the avoidance of iteration over zero-valued blocks. The latter case occurs with $q=4$ and $nf \in [0, 1]$. Running the same experiments with $T_H = L3_{size}$ resulted in a considerable improvement when $nf = 0$, with \texttt{cfO2} becoming even faster than \texttt{tens}, while slow-downs characterized the cases of $nf=1$ for all values of $p$. Our explanation for this is that when $[q=4, nf=0]$, $T_H$ is exceeded by only 40$\%$, while the cost function predicts a save in operation count of 5$\times$. This suggests that our model could be refined to handle the cases in which the gain in operation count is so large that non-extreme variations in working set size should be considered negligible.

\paragraph{Hyperelasticity}
In the hyperelastic model experiments, pre-evaluation is never triggered by \texttt{cfO2}. This is a consequence of the form complexity, which makes the increase factor overwhelm any potential flop reduction. Another distinguishing aspect is the use of vector-valued function spaces, which requires a technique as in Section~\ref{sec:zeros} to avoid wasteful operations over zero-valued entries; \texttt{quad}, \texttt{tens}, \texttt{ufls} and \texttt{cfO2} employ different techniques. Results are displayed in Figure~\ref{fig:hyperelasticity}. \texttt{cfO2} is the best alternative due to removing sharing from the multilinear loop nest and optimization of integration-dependent expressions. \texttt{ufls} performs generally well and exhibits significant speed-ups over non-optimized syntheses; this is a result of the effort in optimizing integration-dependent expressions.

\section{Conclusions}
\label{sec:conclusions}
With this research we have made a first step towards producing a theory and an automated system for the optimal synthesis of loop nests arising in finite element integration. The results are extremely encouraging, suggesting our model applies to a variety of contexts. We have discussed the conditions under which the model only leads to quasi-optimal loop nests. An open problem is understanding how to refine this model to include outer loops. This will probably require exploiting mathematical properties of differential operators. A second open problem is extending our methodology to classes of loops arising in spectral methods; here, the interaction with low level optimization will probably become stronger due to the typically larger working sets deriving from the use of high order function spaces. Lastly, we recall our work is publicly available and is already in use in the latest version of the Firedrake framework.

>>>>

We evaluate the speed-ups achieved by a wide variety of transformation systems over the ``original'' code produced by the FEniCS Form Compiler (i.e., no optimizations applied). We analyze the following transformation systems
\begin{itemize}
\item FEniCS Form Compiler: optimized quadrature mode (work presented in~\citeN{quadrature1}). Referred to as \texttt{quad} 
\item FEniCS Form Compiler: tensor mode (work presented in~\citeN{FFC-TC}). Referred to as \texttt{tens} 
\item FEniCS Form Compiler: automatic mode (choice between \texttt{tens} and \texttt{quad} driven by heuristic, detailed in~\citeN{Fenics} and summarized in Section~\ref{sec:qualitative}). Referred to as \texttt{auto} 
\item UFLACS: a novel back-end for the FEniCS Form Compiler (whose primary goals are improved code generation time and runtime). Referred to as \texttt{ufls} 
\item COFFEE: generalized loop-invariant code motion (work presented in~\citeN{Luporini}). Referred to as \texttt{cfO1} 
\item COFFEE: optimal loop nest synthesis plus symbolic execution for zero-elimination (work of this article). Referred to as \texttt{cfO2} 
\end{itemize}

@@

We report the results of our experiments in Figures~\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity} as three-dimensional plots. The axes represent $q$, $nf$, and code transformation system. We show one subplot for each problem instance $\langle form, nf, q\rangle$, with the code transformation system varying within each subplot. The best variant for each problem instance is given by the tallest bar, which indicates the maximum speed-up over non-transformed code. We note that if a bar or a subplot are missing, then the form compiler failed at generating code because of either exceeding the system memory limit or unable to handle the form.

The rest of the section is structured as follows: we provide insights about the main message of the experimentation; we comment on the impact of autovectorization; we explain in detail, individually for each form, the performance results obtained.

@@

\paragraph{High level view}
The main observation is that our transformation strategy does not always guarantee minimum execution time. In particular, 5$\%$ of the test cases (3 out of 56, without counting marginal differences) show that \texttt{cfO2} was not optimal in terms of runtime. The most significant of such test cases is the elastic model with $[q=4, nf=0]$. There are two reasons for this. Firstly, low level optimization can have a significant impact on actual performance. For example, the aggressive loop unrolling in \texttt{tens} eliminates operations on zeros and reduces the working set size by not storing entire temporaries; on the other hand, preserving the loop structure can maximize the chances of autovectorization. Secondly, memory constraints are critical, particularly the transformation strategy adpopted when exceeding $T_H$. We will later thoroughly elaborate on all these aspects.

\paragraph{Autovectorization}
The discretizations employed result in inner loops and basis function tables of size multiple of the machine vector length. This, combined with the chosen compilation flags, promotes autovectorization in the majority of code variants. An exception is \texttt{quad} due to the presence of indirection arrays in the generated code. In \texttt{tens}, loop nests are fully unrolled, so the standard loop vectorization is not feasible; manual inspection of the compiled code suggests, however, that block vectorization (\cite{SLP-vect}) is often triggered. In \texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces have similar structure (there are a few exceptions in \texttt{cfO2} due to zero-elimination), with loop vectorization being regularly applied, as far as we could evince from compiler reports and manual inspection of assembly code.

\paragraph{Mass}
We start with the simplest of the bilinear forms investigated, the mass matrix. Results are in Figure~\ref{fig:mass}. We first notice that the lack of improvements when $q=1$ is due to the fact that matrix insertion outweighs local assembly. As $q \geq 2$, \texttt{cfO2} generally shows the highest speed-ups. It is worth noting how \texttt{auto} does not always select the fastest implementation: \texttt{auto} always opts for \texttt{tens}, while as $nf \geq 2$ \texttt{quad} would tend to be preferable. On the other hand, \texttt{cfO2} always makes the optimal decision about whether applying pre-evaluation or not.

\paragraph{Helmholtz}
As happened with the mass matrix problem, when $q=1$ matrix insertion still hides the cost of local assembly. For $q \geq 2$, the general trend is that \texttt{cfO2} outperforms the competitors. In particular, with
\begin{itemize}
\item $nf=0$, the adoption of pre-evaluation by \texttt{cfO2} results in increasingly notable speed-ups over \texttt{cfO1}, as $q$ increases; \texttt{tens} is comparable to \texttt{cfO2}, with \texttt{auto} making the right choice. 
\item $nf=1$, \texttt{auto} picks \texttt{tens}; the choice is however sub-optimal when $q=3$ and $q=4$. This can indirectly be inferred from the large gap between \texttt{cfO1/cfO2} and \texttt{tens/auto}: \texttt{cfO2} applies sharing elimination, but it avoids pre-evaluation. 
\item $nf=2$ and $nf=3$, \texttt{auto} reverts to \texttt{quad}, which would theoretically be the right choice (the flop count is much lower than in \texttt{tens} or what would be produced by pre-evaluation); however, the generated code suffers from the presence of indirection arrays, which break autovectorization and ``traditional'' code motion.
\end{itemize}

The sporadic slow-downs or only marginal improvements exhibited by \texttt{ufls} are imputable to the presence of sharing.

An interesting experiment we performed was relaxing the memory threshold by setting it to $T_H = L3_{size}$. We found that this makes \texttt{cfO2} generally slower as $nf \geq 2$, with a maximum slow-down of 2.16$\times$ with $\langle nf=2, q=2\rangle$. The effects of not having a sensible threshold could even be worse in parallel runs, since the L3 cache is shared by the cores.

\paragraph{Elasticity}
The results for the elastic model are displayed in Figure~\ref{fig:elasticity}. The main observation is that \texttt{cfO2} never triggers pre-evaluation, although in some occasions it should. To clarify this, consider the test case $\langle nf=0, q=2 \rangle$, in which \texttt{tens/auto} show a considerable speed-up over \texttt{cfO2}. \texttt{cfO2} finds pre-evaluation profitable -- that is, actually capable of reducing the operation count -- although it does not apply it because otherwise $T_H$ would be exceeded. However, running the same experiments with $T_H = L3_{size}$ resulted in a dramatic improvement, even higher than that of \texttt{tens}. Our explanation is that despite exceeding $T_H$ by roughly 40$\%$, the save in operation count is so large (5$\times$ in this problem) that pre-evaluation would anyway be the optimal choice. This suggests that our model could be refined to handle the cases in which there is a significant gap between potential cache misses and save in flops.

We also note that:
\begin{itemize}
\item the differences between \texttt{cfO2} and \texttt{cfO1} are due to systematic sharing elimination and the use of symbolic execution to avoid iteration over the zero-valued regions in the basis function tables
\item when $nf=1$, \texttt{auto} prefers \texttt{tens} to \texttt{quad}, which leads to sub-optimal operation counts and execution times
\item \texttt{ufls} generally shows better runtime behaviour \texttt{quad} and \texttt{tens}. This is due to multiple facts, including avoidance of indirection arrays, preservation of loop structure, a more effective code motion.
\end{itemize}

\paragraph{Hyperelasticity}
In the experiments on the hyperelastic model, which results are shown in Figure~\ref{fig:hyperelasticity}, \texttt{cfO2} manifests the largest gains out of all problem instances considered in this paper. This is a positive aspect: it means that our transformation algorithm scales with form complexity.

% This is a consequence of the form complexity, which makes the increase factor overwhelm any potential flop reduction. Another distinguishing aspect is the use of vector-valued function spaces, which requires a technique as in Section~\ref{sec:zeros} to avoid wasteful operations over zero-valued entries; \texttt{quad}, \texttt{tens}, \texttt{ufls} and \texttt{cfO2} employ different techniques. Results are displayed in Figure~\ref{fig:hyperelasticity}. \texttt{cfO2} is the best alternative due to removing sharing from the multilinear loop nest and optimization of integration-dependent expressions. \texttt{ufls} performs generally well and exhibits significant speed-ups over non-optimized syntheses; this is a result of the effort in optimizing integration-dependent expressions.

\section{Conclusions}
\label{sec:conclusions}
With this research we have made an important step towards producing a theory and an automated system for the optimal synthesis of loop nests arising in finite element integration. The results are extremely encouraging, suggesting our model applies to a variety of contexts. We have discussed the conditions under which the model only leads to quasi-optimal loop nests. An open problem is understanding how to refine this model to include outer loops. This will probably require exploiting mathematical properties of differential operators. A second open problem is extending our methodology to classes of loops arising in spectral methods; here, the interaction with low level optimization will probably become stronger due to the typically larger working sets deriving from the use of high order function spaces. Lastly, we recall our work is publicly available and is already in use in the latest version of the Firedrake framework.
